---
title: "p4"
output: html_document
---



```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
echo=FALSE, warning=FALSE, message=FALSE)
```




```{r, echo=FALSE}

## Setup

# setwd('<path to folder>/p4')
# install.packages("ggplot2")
# install.packages("knitr", dependencies = T)
# install.packages("dplyr", dependencies = T)
# install.packages("GGally", dependencies = T)
# library(ggplot2)
# library

#getwd()
setwd('/Users/Morgan/Github/udacity-data-analyst-nanodegree/p4')

library(ggplot2)
library(GGally)
library(grid)
library(gridExtra)


## Load the data:

df <- read.csv('wineQualityReds.csv')

```

# Introduction

## About the data

This tidy data set contains 1,599 red wines with 11 variables on the chemical properties of the wine. At least 3 wine experts rated the quality of each wine, providing a rating between 0 (very bad) and 10 (very excellent).

## Guiding question

- Which chemical properties influence the quality of red wines?

## Initial observations

First we'll check out the structure and schema of the data set:

```{r}
str(df)
summary(df)
```


*Note* X is just the id or row number.

It's interesting to noe the min and max values for some of the chemical properties. Density is between .9901 and 1.0037, while total sulfur dioxide ranges from 6 to 289.

We are primarily interested in wine quality, which ranges from 3 - 8. The median is 6 and the average quality rating is 5.636.


# Plots

Let's do some plots!


## Univariate Analysis


```{r echo=FALSE}

# Function that get a basic histogram with a default binwidth of range/30
get_basic_histogram <- 
  function(column, data, binwidth = diff(range(data[column]))/30) {
    return(ggplot(aes_string(x = column), data = data) + 
             geom_histogram(binwidth = binwidth) + 
              ggtitle(paste('Histogram of',column))

           )
  }


```

### Fixed Acidity


```{r }

get_basic_histogram("fixed.acidity", df)
summary(df['fixed.acidity'])

```

It looks like the distribution of fixed acidity is skewed positive. But there don't appear to be any outliers. Let's check count of quality against fixed acidity...

```{r }

ggplot(aes(x = factor(df$quality), y = fixed.acidity), data = df) +
  geom_bar(stat = 'summary', fun.y = mean)

```

These bar charts show the average fixed acidity levels per quality. There seems to be a vague positive correlation, except at the beginning and end. Later on, I'd like to compare ratings with some combination of acidity measurements...


### Volatile Acidity

```{r }
get_basic_histogram("volatile.acidity", df, 0.1)
summary(df['volatile.acidity'])

```

Volatile acidity is also skewed positive by a relatively small number of outliers. Let's remove them.

```{r }
get_basic_histogram("volatile.acidity", df, 0.1) + xlim(0,1.4)
summary(df['volatile.acidity'])

```

The media and mean are both around .5, while the max is around 1.58. Let's compare this type of acidity against quality.

```{r }

ggplot(aes(x = factor(df$quality), y = volatile.acidity), data = df) +
  geom_bar(stat = 'summary', fun.y = mean)

```

Definitely some correlation here! Looks like bad tasting wines have high levels of volatile acidity. We'll have to explore this further later...

### Citric Acid

```{r }
get_basic_histogram("citric.acid", df)
summary(df['citric.acid'])

```

This does not resemble a normal distribution. Apparently, very many wines have little to no citric acid at all. I suppose rice wines would fall into this category. It would be interesting to know if some grape wines can be made without any citric acid. Let's see how citric acid levels look against quality.


```{r }

ggplot(aes(x = quality, y = citric.acid), data = df) +
  geom_bar(stat = 'summary', fun.y = mean)

```

Ok that seems very positively correlated. Right now, volatile acidity and citric acid are the principle predictor variables. Such as, Y = aX1 + bX2. Where Y is quality, X1 is volatile acidity and X2 is citric acid. Before we go looking for other predictor variables, let's just make a grid of histograms for the remaining characteristics.


```{r }
grid.arrange(
  get_basic_histogram("residual.sugar", df),
  get_basic_histogram("chlorides", df),
  get_basic_histogram("free.sulfur.dioxide", df),
  get_basic_histogram("total.sulfur.dioxide", df),
  get_basic_histogram("density", df),
  get_basic_histogram("pH", df),
  get_basic_histogram("sulphates", df),
  get_basic_histogram("alcohol", df)

)

```

Ok so the majority of these remaining distributions skew right. That makes sense, you can't have negative residual sugar, you can't have negative chlorides, etc. There's always a minimum and never a maximum. I'm going to limit some of the distributions to remove outliers and see how they look then.

```{r }
grid.arrange(
  get_basic_histogram("residual.sugar", df) + xlim(0,10),
  get_basic_histogram("chlorides", df) + xlim(0,.3),
  get_basic_histogram("free.sulfur.dioxide", df) + xlim(0, 60),
  get_basic_histogram("total.sulfur.dioxide", df) + xlim(0, 200),
  get_basic_histogram("density", df),
  get_basic_histogram("pH", df),
  get_basic_histogram("sulphates", df) + xlim(0, 1.75),
  get_basic_histogram("alcohol", df)
)

```

The histograms for chlorides and sulphates look a lot more gaussian after doing that! Before we move on to correlations and multivariate analysis, let's look at distribution of quality!


### Distribution of Quality


```{r }
ggplot(aes_string(x = factor(df$quality)), data = df) +
             geom_bar(stat="count") + xlab('Quality')
summary(df['quality'])

```

It appears to be a normal distribution of quality.


The "quality" of each wine is one of 6 discrete values: 4, 5, 6, 7, 8.
The distribution of wines by quality appears to be a bell curve - most wines are in the middle of the range, with quality ratings of 5 and 6.




### Correlation matrix

Let's get a quick look at the correlation between all variable pairs


```{r }
ggcorr(df)
```

Looks like strong positive correlations between quality and alcohol. Fairly significant positive correlation between quality and sulphates (We will explore this later), as well as citric acid. And strong negative correlations between quality and volatile.acidity; we saw an indication of that in an earlier bar chart. If I had to come up with predictor variables right now, I'd say these are them:

-alcohol
-sulphates
-citric.acid
-volatile.acidity

... and maybe fixed acidity.


## Bivariate Analysis


### Quality by Alcohol

Now, let's look at some of the variables that appear to have a linear relationship with quality.

``` {r Quality by Alcohol Scatterplot}
ggplot(data = df,
       aes(x = alcohol, y = quality)) +
  geom_point() + 
  scale_x_continuous() + 
  scale_y_continuous() +
  ggtitle('Quality by Alcohol')
```

Looks like overplotting. So let's add some jitter and transparency to the points:

``` {r Quality by Alcohol with Jitter and Transparency}
ggplot(data = df,
       aes(x = alcohol, y = quality)) +
  geom_point(position = 'jitter', alpha = 0.5, size = 0.75) + 
  scale_x_continuous() + 
  scale_y_continuous() +
  ggtitle('Quality by Alcohol with Jitter and Transparency')
```

Now let's throw a regression line in there!

``` {r Quality by Alcohol with Jitter and Transparency with a Regression}
ggplot(data = df,
       aes(x = alcohol, y = quality)) +
  geom_point(position = 'jitter', alpha = 0.5, size = 0.75) + 
  scale_x_continuous() + 
  scale_y_continuous() +
  ggtitle('Quality by Alcohol with Jitter and Transparency and a Regression') +
  geom_smooth(method='lm',formula=y~x)
```


Well it's a positive correlation alright. Albeit, It probably has a low R squared value, judging by how scattered the points are. Let's see:

``` {r}

# ?lm
library(memisc)

m1 <- lm(quality ~ alcohol, data = df)
mtable(m1, sdigits = 4)
?mtable
```

Ok, we have a starting point. R-squared is .2267 with Alcohol as our only predictor variable. Let's continue...

### Box plots


Thanks to feedback from the reviewer of my first submission, let's try a box plot instead of a scatter plot.

``` {r Quality by Alcohol}
ggplot(data = df,
       aes(x = factor(quality), y = alcohol)) +
  geom_boxplot() + 
  ggtitle('Quality by Alcohol')
```

Look at all the outliers for 5s... We know that there were mor 5s than any other rating. This may have been the default for reviewers who just didn't feel any particular way. I certainly wouldn't exclude any of these data points, they look legitimate.

#### Quality by Volatile Acidity

There might still be a linear relationship here

``` {r Quality by Volatile Acidity with Jitter and Transparency}
ggplot(data = df,
       aes(x = factor(quality), y = volatile.acidity)) +
  geom_boxplot() + 
  ggtitle('Quality by Volatile Acidity with Jitter and Transparency')
```

Just another illustration of a negative correlation with quality: the lower the volatile acidity, the better the wine. There are outliers but I wouldn't worry about them.


#### Quality by Sulphates

``` {r Quality by Sulphates with Jitter and Transparency}
ggplot(data = df,
       aes(x = factor(quality), y = sulphates)) +
  geom_boxplot() + 
  ggtitle('Quality by Sulphates with Jitter and Transparency')
```

Lots of outliers, weak but positive correlation. This is going to have a lower R-squared value than alcohol I imagine... well obviously, the correlation matrix pointed that out already. But let's see just how much less the R-squared value is...

``` {r}

m1 <- lm(quality ~ sulphates, data = df)
mtable(m1, sdigits = 4)
?mtable
```

An R-squared value of .0632 is pretty low. I doubt this, combined with alcohol as a predictor variable, is going to improve the combined R-squared value very much.



### New Variable

I created a new variable "rating" has 3 values: bad, average, good

``` {r}
df$rating <- ifelse(df$quality < 5, 'bad', ifelse(
  df$quality < 7, 'average', 'good'))
df$rating <- ordered(df$rating,
                     levels = c('bad', 'average', 'good'))
summary(df$rating)
```



### Grid of boxplots with variables against new variable Rating

Now let's recreate those box plots with the rating instead of the quality variable. I'm going to look at density, since I usually feel like I favor a lighter wine. But then again, who's to say what I think is "light" is less dense...


``` {r Boxplots against ratings}

p1 <- ggplot(data = df,
     aes(x = rating, y = alcohol)) +
  geom_boxplot() + 
  ggtitle('Rating by Alcohol')

p2 <- ggplot(data = df,
       aes(x = rating, y = density)) +
  geom_boxplot() + 
  ggtitle('Rating by Density')


p3 <- ggplot(data = df,
       aes(x = rating, y = volatile.acidity)) +
  geom_boxplot() + 
  ggtitle('Rating by Volatile Acidity')


p4 <- ggplot(data = df,
       aes(x = rating, y = sulphates)) +
  geom_boxplot() + 
  ggtitle('Rating by Sulphates')

grid.arrange(p1, p2, p3, p4)

```

No big suprises here. Stronger correlations for alcohol and volatile acidity and sulphates. Very little to learn from the Density box plot - outliers in the middle quality category and little correlation.


### Multivariate Analysis


Now let's look at density and alcohol against ratings:

```{r}

model <- lm(density ~ alcohol + rating, data=df)
grid <- with(df, expand.grid(
  alcohol = seq(min(alcohol), max(alcohol)),
  rating = levels(rating)
))

grid$density <- stats::predict(model, newdata=grid)


ggplot(data = df,
       aes(y = density, x = alcohol)) +
  geom_point(aes(color = rating)) + 
  scale_colour_brewer(palette = 'Set1') +
  geom_line(data=grid)


```

#### Note: Added a regression line after review of prior submission

It seems that as alcohol content increases, density decreases. Perhaps alcohol, by nature is more dense than the ingredients of wine? This last plot also indicates that, holding alcohol constant, density does not affect quality.


``` {r}


model <- lm(sulphates ~ alcohol + rating, data=df)
grid <- with(df, expand.grid(
  alcohol = seq(min(alcohol), max(alcohol)),
  rating = levels(rating)
))

grid$sulphates <- stats::predict(model, newdata=grid)

ggplot(data = df,
       aes(y = sulphates, x = floor(alcohol),
           color = rating)) +
  geom_point(position = 'jitter', alpha = 0.5, size = 0.75) + 
  scale_colour_brewer(palette = 'Set1') +
  scale_y_continuous(limits=c(0.3,1.5)) +
  scale_fill_gradient(low="green",high="darkgreen") +
    geom_line(data=grid)


```

It looks like that holding alcohol constant, then more sulphates improves the quality of the wine.



## Final Plots


### Plot one

``` {r}


model <- lm(sulphates ~ alcohol + rating, data=df)
grid <- with(df, expand.grid(
  alcohol = seq(min(alcohol), max(alcohol)),
  rating = levels(rating)
))

grid$sulphates <- stats::predict(model, newdata=grid)

ggplot(data = df,
       aes(y = sulphates, x = alcohol,
           color = rating)) +
  geom_point() +
  scale_y_continuous(limits=c(0.3,1.5)) +
  ylab("potassium sulphate (g/dm3)") +
  xlab("alcohol (% by volume)") +
  scale_colour_brewer(palette = 'Set1') +
  geom_line(data=grid) + 
  ggtitle("Alcohol and sulphates over wine quality")

```

What I did learn is that while there may be a correlation between alcohol and quality, or sulphates and quality (see correlation matrix above), there is little or no correlation between alcohol and sulphates.


## Plot two: Quality by Acidic Attributes

``` {r}
grid.arrange(ggplot(data = df, aes(x = factor(quality), y = fixed.acidity)) + 
               ylab('Fixed Acidity') +
               xlab('Quality') +
               geom_boxplot(),
             ggplot(data = df, aes(x = factor(quality), y = volatile.acidity)) +
               ylab('Volatile Acidity') +
               xlab('Quality') +
               geom_boxplot(), 
             ggplot(data = df, aes(x = factor(quality), y = citric.acid)) +
               ylab('Citric Acid') +
               xlab('Quality') +
               geom_boxplot(), 
             ggplot(data = df, aes(x = factor(quality), y = pH)) +
               ylab('pH') +
               xlab('Quality') +
               geom_boxplot())

```


This shows that there is a stronger correlation between quality and volatile acidity and citric acid then there is between quality and fixed acidity and pH.


## Plot three: Model


```{r, echo=FALSE}

# ?lm
library(memisc)

```

``` {r}


m1 <- lm(quality ~ alcohol, data = df)
m2 <- update(m1, ~ . + sulphates)
m3 <- update(m2, ~ . + citric.acid)
m4 <- update(m3, ~ . + volatile.acidity)
m5 <- update(m4, ~ . + fixed.acidity)
mtable(m1, m2, m3, m4, m5)


```

The R squared value of m5 is 0.3. Which means that our regression model only explains around 30% of the variance in quality. It seems we are missing the information that would better predict the quality of wine.

```{r, echo=FALSE}

# ?lm
# install.packages("memisc")
library(dplyr)

```

``` {r}

set.seed(87654321)
training_data <- sample_frac(df, .6)
test_data <- df[ !df$X %in% training_data$X, ]

results <- data.frame(
  test_data$quality,
  predict(m5, test_data) - as.numeric(test_data$quality)
)
names(results) <- c("quality", "error")
ggplot(data=results, aes(x=quality,y=error)) +
  geom_point(position = 'jitter', alpha = 0.5, size = 0.75)
```

The residual plot indicates wide swings in error.

And let's see it as a box plot...

``` {r}
ggplot(data = results,
       aes(y = error, x = quality, group = factor(quality))) +
  geom_boxplot()

```

This IS interesting. The error is so much larger for poorly rated wines. While at the other end of the spectrum, the error is much lower...

Anyway, judging by the regression model, there must be some key information missing that might better predict quality.



## Reflection

From looking at the initial ggpairs plot, it did seem that there was some correlation between quality and a couple variables provided by the data set, namely: alcohol, sulphates, citric.acid, fixed.acidity, and volatile.acidity. Density may have been another variable we could have used in our model.

From looking at the distributions of chemical variables, many had ver long tails and skewed distributions. Only a couple had normal distributions (density, pH).

For many variables, a scatterplot against quality did not indicate much. Even after adding jitter to account for overplotting.

For future studies, it would be helpful to have any additional ingredients. Also, if it were possible to segment the wines, we could hold constant certain qualities - I'm no wine expert, but I believe cabernets probably have different chemicals than merlots, etc.

Another consideration: we cannot account for taste. We do not know who the tasters are. Certainly, different people taste things differently, and that may account for varying quality ratings.

What I did learn is that wines with high alcohol and sulphates do tend to be rated higher. So, those are ingredients I will look for when making my next wine purchase.

### Difficulties
The hardest part was understanding the language of ggplot. For example, trying to build a histogram of quality using factor('quality') will throw an error and R studio will advise "perhaps you want to use stat='count'". But, there's no suggestion of where to put it. After googling it for 10 minutes, you find that it is meant to be used with geom_bar() instead. But, understandably, there is a syntax and it's documented and it was at first esoteric but I'm slowly getting used to it!




